Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2019-11-05 manuscript/climate_sensitivity_draft.Rmd
update summary and results

text | word count | other | n
---------------------------------------------|---------------------|-----------------------------|--------------------
Total word count (excluding summary, references and legends) |  currently ~6442 (strict limit 6,500)	 |	No. of figures | 	2 (both colour)
Summary|  	currently 198 (limit 200) | 	No. of Tables | 	5  
Introduction|	currently ~1339 |	No of Supporting Information files | 	###.
Materials and Methods |	currently ~1604	||	 
Results |	currently ~1558	 ||
Discussion |	currently ~1866 (limit 30% of total (not strict), or 1950 if manuscript reaches word limit)	||
Acknowledgements | currently ~73	||

@@

- Individual-level drought resistance decreased with tree height, whereas crown position explained little additional variation and root water access did not confer additional resistance. *Resistance was lower among diffuse-porous species and those whose leaves shrunk more upon dehydration and lost turgor at less negative leaf water potential.* However, there was substantial variation in the best predictor variables across given droughts.

@@

*Key words*:  canopy position; drought; Forest Global Earth Observatory (ForestGEO); hydraulic traits; temperate broadleaf deciduous forest;  tree growth; tree height; tree-ring [5-8] *Five to eight key words (in alphabetical order) . Words that are in the title can, and should, be among these. Very short phrases and scientific names with their common equivalents (e.g. Nicotiana tabacum (tobacco)) are acceptable.*

@@

Here, we combine tree-ring records covering three droughts (1966, 1977, 1999), species functional and hydraulic trait measurements, and forest census data from a 25.6-ha ForestGEO plot in Virginia (USA) to test a series of hypotheses and associated specific predictions (Table 1) designed to yield functional understanding of how tree size, microenvironment, and species' traits collectively shape drought responses. 
First, we focus on the role of tree size and its interaction with microenvironment. 
We confirm that, consistent with most forests globally, larger-diameter trees tend to have lower drought resistance in this forest, which is in an ecoregion represented by only one study in [@bennett_larger_2015] (*H1.0*). 
We then test hypotheses designed to disentangle the relative importance of tree height (*H1.1*), crown exposure (*H1.2*), and root water access, which should be greater for larger trees in dry but not in perpetually wet microsites (*H1.3*). 
Second, we focus on the role of species' functional and hydraulic traits and their interaction with tree height. We hypothesize that drought resistance will follow observed patterns in relation to wood density (negative effect; @hoffmann_hydraulic_2011) and specific leaf area (positive effect), but that hydraulic traits including xylem architecture (*i.e.*, ring, semi-ring, or diffuse porous), leaf area shrinkage upon dehydration, and turgor loss point will prove better predictors (*H2.1*). 
We then test whether these traits correlate with tree height (*H2.2*), potentially driving the observed tendency for taller trees to suffer more during drought (*H2.3*). 
Finally, we focused on variability among droughts, asking how community resistance varied across droughts (*H3.1*) and whether the factors confirming vulnerability or resistance varied across droughts (*H3.2*).

@@

We partially support *H2.1*: Species' hydraulic traits --$XP$, $PLA_{dry}$, and $\pi_{tlp}$--were sometimes useful in explaining variation in drought responses, whereas $LMA$ and $WD$ were not (Tables 1,4,5). Specifically, $LMA$ and $WD$ never significantly associated with $R$ in the univariate models (all dAIC $\le$ 0.22; Table 4), and therefore these were excluded as candidate variables for the full multivariate models. In contrast, $XP$, $PLA_{dry}$, and $\pi_{tlp}$ all explained at least modest amounts of variation (dAIC > 1.0) in at least one drought (Table 4). Of these, $PLA_{dry}$ was the strongest predictor, with consistently negative coefficients across all droughts. $\pi_{tlp}$ never came out as significant (dAIC $\ge$ 2) in the univariate models, but had a consistently negative coefficient (Table 4). Whereas ring-porous species had highest $R$ overall and in the 1966 and 1999 droughts, they had lower $R$ in 1977. Results were similar in the context of multivariate models (Table 5), except that $\pi_{tlp}$ had a positive coefficient in the 1966 models in which it was included.

@@

We mostly reject the hyopthesis (*H3.2*) that directions of responses varied across droughts. In the majority of cases, response directions were consistent across droughts in both univariate and multivariate models (Tables 1,4,5). However, there were a few exceptions--most commonly in the categorical variables ($CP$ and $XP$) but also for $\pi_{tlp}$ in the multivariate model for the 1966 drought; Tables 4, 5). These differences may very well be random, as opposed to statistically meaningful. Among the univariate models, there was no instance where predictor variables signficantly improved the models of two different droughts (dAIC $\ge$ 2), but with contrasting coefficients (Table 4). Among the multivariate models, $CP$ was not consistently in the top models for any drought (Table 5), and $\pi_{tlp}$ only appeared with a positive coefficient in two of five models for the 1966 drought (contrasting with a negative coefficient in the univariate model; Table 4). The difference most likely to be real is that ring and semi-ring porous species had lower resistance than diffuse porous species in the 1977 drought, contrasting with higher resistance in 1966 and 1999 (Tables 4,5), but note that $XP$ was not a signficant predictor on its own for the 1977 drought.

We support the hypothesis (*H3.3*) that the strength of predictor variables was different across the droughts (Tables 1,4,5). For instance, $ln[H]$ and $PLA_{dry}$ had much stronger negative effects in 1966 than in the other two years, $ln[TWI]$ had the strongest negative effect in 1977, and $CP$ (lowest $R$ among supressed trees) and $XP$ (lowest $R$ among diffuse-porous trees) were strongest in 1999 (Tables 4,5).

>>>>

text | word count | other | n
---------------------------------------------|---------------------|-----------------------------|--------------------
Total word count (excluding summary, references and legends) |  currently ~6442 (strict limit 6,500)	 |	No. of figures | 	2 (both colour)
Summary|  	currently 200 (limit 200) | 	No. of Tables | 	5  
Introduction|	currently ~1339 |	No of Supporting Information files | 	###.
Materials and Methods |	currently ~1604	||	 
Results |	currently ~1558	 ||
Discussion |	currently ~1866 (limit 30% of total (not strict), or 1950 if manuscript reaches word limit)	||
Acknowledgements | currently ~73	||

@@

- Individual-level drought resistance decreased with tree height, whereas crown position explained little additional variation and root water access did not confer additional resistance. Resistance was lower among species whose leaves shrunk more upon dehydration and lost turgor at less negative leaf water potential. However, there was substantial variation in the best predictor variables across given droughts.

@@

*Key words*:  canopy position; drought; Forest Global Earth Observatory (ForestGEO); leaf hydraulic traits; temperate broadleaf deciduous forest;  tree growth; tree height; tree-ring [5-8] *Five to eight key words (in alphabetical order) . Words that are in the title can, and should, be among these. Very short phrases and scientific names with their common equivalents (e.g. Nicotiana tabacum (tobacco)) are acceptable.*

@@

Here, we combine tree-ring records covering three droughts (1966, 1977, 1999), species functional and hydraulic trait measurements, and forest census data from a 25.6-ha ForestGEO plot in Virginia (USA) to test a series of hypotheses and associated specific predictions (Table 1) designed to yield functional understanding of how tree size, microenvironment, and species' traits collectively shape drought responses. 
First, we focus on the role of tree size and its interaction with microenvironment. 
We confirm that, consistent with most forests globally, larger-diameter trees tend to have lower drought resistance in this forest, which is in an ecoregion represented by only one study in [@bennett_larger_2015] (*H1.0*). 
We then test hypotheses designed to disentangle the relative importance of tree height (*H1.1*), crown exposure (*H1.2*), and root water access, which should be greater for larger trees in dry but not in perpetually wet microsites (*H1.3*). 
Second, we focus on the role of species' functional and hydraulic traits and their interaction with tree height. We hypothesize that drought resistance will follow observed patterns in relation to wood density (negative effect; @hoffmann_hydraulic_2011) and specific leaf area (positive effect), but that hydraulic traits including xylem architecture (*i.e.*, ring or diffuse/ semi-ring porous), leaf area shrinkage upon dehydration, and turgor loss point will prove better predictors (*H2.1*). 
We then test whether these traits correlate with tree height (*H2.2*), potentially driving the observed tendency for taller trees to suffer more during drought (*H2.3*). 
Finally, we focused on variability among droughts, asking how community resistance varied across droughts (*H3.1*) and whether the factors confirming vulnerability or resistance varied across droughts (*H3.2*).

@@

We partially support *H2.1*: Species' hydraulic traits --$XP$, $PLA_{dry}$, and $\pi_{tlp}$--were sometimes useful in explaining variation in drought responses, whereas $LMA$ and $WD$ were not (Tables 1,4,5). Specifically, $LMA$ and $WD$ never significantly associated with $R$ in the univariate models (all dAIC $\le$ 0.22; Table 4), and therefore these were excluded as candidate variables for the full multivariate models. In contrast, $XP$, $PLA_{dry}$, and $\pi_{tlp}$ all explained at least modest amounts of variation (dAIC > 1.0) in at least one drought (Table 4). Of these, $PLA_{dry}$ was a strong predictor for the 1966 drought and overall, with consistently negative coefficients across all droughts. $\pi_{tlp}$ never came out as significant (dAIC $\ge$ 2) in the univariate models, but had a consistently negative coefficient (Table 4). Whereas ring-porous species had highest $R$ overall and in the 1966 and 1999 droughts, they had lower $R$ in 1977. Results were similar in the context of multivariate models (Table 5), except that $PLA_{dry}$ had a small positive coefficient in the one 1999 model in which it was included.

@@

We mostly reject the hyopthesis (*H3.2*) that directions of responses varied across droughts. In the majority of cases, response directions were consistent across droughts in both univariate and multivariate models (Tables 1,4,5). However, there were a few exceptions--most commonly in the categorical variables ($CP$ and $XP$) but also for $PLA_{dry}$ in one multivariate model for the 1999 drought (Tables 4, 5). These differences may very well be random, as opposed to statistically meaningful. Among the univariate models, there was no instance where predictor variables signficantly improved the models of two different droughts (dAIC $\ge$ 2), but with contrasting coefficients (Table 4). Among the multivariate models, $CP$ was not consistently in the top models for any drought (Table 5), and $PLA_{dry}$ only appeared with a positive coefficient in the weakest of six models for the 1999 drought (contrasting with a negative coefficient in the univariate model; Table 4). The difference most likely to be real is that ring-porous species had lower $R$ than diffuse- and semi-ring- porous species in the 1977 drought, contrasting with higher resistance in 1966 and 1999 (Tables 4,5), but note that $XP$ was not a signficant predictor on its own for the 1977 drought.

We support the hypothesis (*H3.3*) that the strength of predictor variables was different across the droughts (Tables 1,4,5). For instance, $ln[H]$ and $PLA_{dry}$ had much stronger negative effects in 1966 than in the other two years, $ln[TWI]$ had the strongest negative effect in 1977, and $XP$ (higher $R$ among ring-porous trees) was strongest in 1999 (Tables 4,5).
